---
title: "20240202 gDT panel 1 MNP workspace FlowWorkspace proportion analysis"
author: "Abbey Figliomeni"
date: "2024-02-02"
output: html_document
---

# Data import and preprocessing
## Extract flow data (class: gatingset) using cytoML
```{r, results="hide"}
source("[cytoML script working directory]")

```

## Check data
```{r}
## Check data
### visualize gating nodes
gs_get_pop_paths(gs, path="auto")
plot(gs)

```

# FLOW WORKSPACE: Creating dataframe with cell subset proportions and TF ratios
```{r}
flow_df <- pData(gs)  # creating df based on pData (includes all clinical variables)
flow_df$Panel <- "gDT1"  # append panel to pData

sample_list <- row.names(flow_df)  # row names are IDs
sample_list

pop_list <- gs_get_pop_paths(gs, path="auto")  # list of gating nodes populations from flowjo
pop_list
pop_list <- pop_list[!grepl("Q[0-9]\\:", pop_list)] # remove unnamed quad gates
pop_list

## compute surface marker percentage of parent ----
for (s in sample_list){
  for (p in pop_list){
    stat <- gh_pop_get_stats(gs[[s]], nodes=p, type="percent")
    flow_df[s, p] <- stat$percent*100
  }
}

## compute relevant marker MFI ----
MFIlist <- c("CXCR4 BV421", "CD39 BV605",  "CD16 BV711", "CD25 BV785", "HLA-DR Kblue", "CD127 PE-Dazzle595", "CD33 PE", "CD14 PE-Cy5", "CD11b BV510")

for (s in sample_list){
  for (m in MFIlist){
    Mono <- gh_pop_get_stats(gs[[s]], nodes ="NotMono", type=pop.MFI, inverse.transform=T)
    y <- paste("MonoAll", m , "MFI")
    flow_df[s, y] <- Mono[[m]]
    
    cMono <- gh_pop_get_stats(gs[[s]], nodes ="cMono", type=pop.MFI, inverse.transform=T)
    y <- paste("cMono", m , "MFI")
    flow_df[s, y] <- cMono[[m]]
    
    iMono <- gh_pop_get_stats(gs[[s]], nodes ="iMono", type=pop.MFI, inverse.transform=T)
    y <- paste("iMono", m , "MFI")
    flow_df[s, y] <- iMono[[m]]
    
    ncMono <- gh_pop_get_stats(gs[[s]], nodes ="ncMono", type=pop.MFI, inverse.transform=T)
    y <- paste("ncMono", m , "MFI")
    flow_df[s, y] <- ncMono[[m]]
    
    mMDSC <- gh_pop_get_stats(gs[[s]], nodes ="M-MDSCs", type=pop.MFI, inverse.transform=T)
    y <- paste("mMDSC", m , "MFI")
    flow_df[s, y] <- mMDSC[[m]]
  }
}

```

# Adding myeloid proportions
* monocyte subsets already as proportions of total monocytes, however need to convert "monocytes all" and "M-MDSC" gates to as percentage of myeloid, rather than as % of HLA-DR+ or HLA-DR-
```{r}
# converting to percentage of myeloid CD33+ cells
flow_df$MoncytesAll_pcMyeloid <- flow_df$MonocytesAll * flow_df$`HLA-DR+`/100 # monocytes
flow_df$mono_sum_pcMyeloid <- flow_df$NotMono * flow_df$`HLA-DR+`/100  # sum subsets
flow_df$`M-MDSC_pcMyeloid` <- flow_df$`M-MDSCs` * flow_df$`HLA-DRneg`/100 # M-MDSCs

# percentage of PBMCs
flow_df$MoncytesAll_pcPBMCs <- flow_df$MoncytesAll * flow_df$`HLA-DR+`/100 * flow_df$`Myeloid cells`/100 * flow_df$LinNeg/100 # monocytes
flow_df$mono_sum_pcPBMCs <- flow_df$mono_sum_pcMyeloid * flow_df$`Myeloid cells`/100 * flow_df$LinNeg/100
flow_df$`M-MDSC_pcPBMCs` <- flow_df$`M-MDSCs` * flow_df$`HLA-DRneg`/100 * flow_df$`Myeloid cells`/100 * flow_df$LinNeg/100 # monocytes

# total MDSCs
flow_df$MDSCall_pcMyeloid <- (flow_df$`M-MDSCs` + flow_df$`MDSC other`) * flow_df$`HLA-DRneg`/100 # M-MDSCs

```



## Edit dataframe column names
```{r}
## Editing colnames of flow DF  ----
colnames(flow_df)
flow_df$root <- NULL
flow_df$LinNeg <- NULL
colnames(flow_df) <- gsub("\\/", " ", colnames(flow_df))  # substitutes all slashes in node names for spaces
colnames(flow_df) <- gsub("\\+", "pos", colnames(flow_df))  # substitutes all pluses in node names for "pos"
colnames(flow_df) <- gsub(" ", "_", colnames(flow_df))  # substitutes all spaces in node names for underscores
colnames(flow_df) <- gsub("\\-", ".", colnames(flow_df))  # substitutes all dashes in node names for dots
colnames(flow_df)
```


## Save as CSV
```{r, warning=F}
flow_df <- tibble::rownames_to_column(flow_df, "ID")

write.csv(flow_df, "[working directory]", row.names = F)
```


# Statistical analysis
Staistical workflow:
* test PD v control for all basic flow metrics
* IF PD v Control SIGNIFICANT (p < 0.05, no trends), THEN:
    + correlate with continuous HY and UPDRS scales
    + correlate with disease duration, GSRS score, dietary scales???
    + make publication-ready boxplot figure (also do for important non-sig stats such as overall gDT)
    
**No** subgrouping as confounds interpretation and doubles up with correlations.

For significant correlations, look into constructing GLMM.


## Libraries
```{r, echo=F, message=F}
library(rstatix)
library(dplyr)
library(ggplot2)
library(psych)
library(car)
library(Hmisc)
library(corrplot)
```


## Prepare data
```{r}
str(flow_df)
flow_df$Condition <- factor(flow_df$Condition, levels=c("Control", "PD"))
flow_df$DiseaseDuration <- as.numeric(flow_df$DiseaseDuration)
flow_df$LEDD <- as.numeric(flow_df$LEDD)
flow_df$Age <- as.numeric(flow_df$Age)
flow_df$GSRStotal <- as.numeric(flow_df$GSRStotal)
flow_df$H_and_Y <- as.numeric(flow_df$H_and_Y)
flow_df$UPDRS3 <- as.numeric(flow_df$UPDRS3)
flow_df$AAO <- as.numeric(flow_df$AAO)



colnames(flow_df)
gDT_1_stat_list <- colnames(flow_df)[14:length(colnames(flow_df))]  # colnames of columns with flow data
gDT_1_stat_list

```

## PD v Control
```{r}
flow_df_PD <- subset(flow_df, Condition == "PD")
flow_df_Ctrl <- subset(flow_df, Condition == "Control")

gDT_1_stats <- data.frame(Parameter = NA,  # create empty data frame to deposit stats into below
                           StatTest=NA,
                           pValue=NA,
                           med.mean.PD=NA,
                           med.mean.HC=NA,
                           med.mean.FC=NA)


for(i in gDT_1_stat_list){
  pos <- which(gDT_1_stat_list %in% i)
  gDT_1_stats[pos, "Parameter"] <- i
  SW_test_PD <- shapiro_test(flow_df_PD, i)
  SW_test_C <- shapiro_test(flow_df_Ctrl, i)
  if(SW_test_PD$p<0.05 | SW_test_C$p<0.05){
    MW <- wilcox.test(flow_df_PD[[i]], flow_df_Ctrl[[i]])
    gDT_1_stats[pos,"StatTest"] <- "MannWhitneyU"
    gDT_1_stats[pos, "pValue"] <- MW$p.value
    gDT_1_stats[pos, "med.mean.HC"] <- median(flow_df_Ctrl[[i]])
    gDT_1_stats[pos, "med.mean.PD"] <- median(flow_df_PD[[i]])
    gDT_1_stats[pos, "med.mean.FC"] <- median(flow_df_PD[[i]])/median(flow_df_Ctrl[[i]])
  } else{
    TT <- t.test(flow_df_PD[[i]], flow_df_Ctrl[[i]])
    gDT_1_stats[pos, "StatTest"] <- "Ttest"
    gDT_1_stats[pos, "pValue"] <- TT$p.value
    gDT_1_stats[pos, "med.mean.HC"] <- mean(flow_df_Ctrl[[i]])
    gDT_1_stats[pos, "med.mean.PD"] <- mean(flow_df_PD[[i]])
    gDT_1_stats[pos, "med.mean.FC"] <- mean(flow_df_PD[[i]])/mean(flow_df_Ctrl[[i]])
  }
}
```

### Save CSV
```{r, warning=F}
write.csv(gDT_1_stats, "20240203 gDT panel 1 MNP PD v Control stats.csv", row.names = F)
```


# Visualise Data
## PD v Control boxplots 
```{r, warning=F}
# creating function to loop through and plot all flow variables by condition
setwd("[working directory]")

# with points displayed and no box
customPlot <- function(varName) {
  
  p <- flow_df %>% 
    group_by_("Condition") %>% 
    select_("Condition",varName) %>% 
    ggplot(aes_string("Condition",varName)) + 
    stat_boxplot(geom ='errorbar', width=0.2, lwd=1) +
    geom_boxplot(show.legend = FALSE, aes(fill=Condition), lwd=0.8, outlier.shape = NA) +
    geom_jitter(shape=1, position=position_jitter(0.2, seed = 1), size=4) +
    geom_label( 
    data=flow_df, # Filter data first
    position=position_jitter(0.2, seed = 1), aes(label=ID)) +
    scale_fill_manual(values = c("mediumslateblue", "orange1")) +
    labs(x="") +
    theme(axis.text.x = element_text(size=14, colour="black"),
          axis.text.y = element_text(size=12),
          axis.title.y = element_text(size=14))
  filename <- paste0(varName, '.jpeg')
  ggplot2::ggsave(plot = p, filename = paste0(filename),
                  width = 2870,
                  height = 2072,
                  units = "px") # width 3.6 default, height 5 default
  
}
lapply(gDT_1_stat_list, customPlot)

```

# Significant data - correlations with pData
"ClinicalSignifFlowCor" function takes inputs: flow dataframe, stats dataframe, and a significance threshold (alpha) and calculates the relevant correlations between clinical variables and any significantly altered flow variables. Output: dataframe containing variables, correlation coefficient (spearman rank), and p-value. NB - ensure relevant clinicial variables are numeric.
```{r}
source("FlowWorkspace significant flow variables correlation with pData function.R")
ClinicalSignifFlowCor(flow_df=flow_df, stats_df=gDT_1_stats, alpha=1)
write.csv(Clinical_correlations, "[working directory]", row.names = F)

```

